---
title: "Lecture 13"
subtitle: "Learning Models"
author: Tyler Ransom
date: ECON 6343, University of Oklahoma
output:
  xaringan::moon_reader:
    includes:
        in_header: "13slides_files/mathjax-equation-numbers.html"
    css: ['default', 'metropolis', 'metropolis-fonts', 'ou-colors.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

---

```{r, load_refs, include=FALSE, cache=FALSE}
library(RefManageR)
library(tidyverse)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           hyperlink = FALSE,
           dashed = TRUE)
biblio <- ReadBib("../../References/References.bib", check = FALSE)
```

# Plan for the Day

1. Learning from noisy signals

2. How learning models relate to factor models

3. Bayesian updating & the Kalman filter

4. How to code learning models

5. Examples of learning models in economics


---
# Attribution

Some of these slides are based on content from Peter Arcidiacono's course on learning models.


---
# Imperfect information

- Imperfect information abounds in economics (and real life)

    - life is full of "noisy signals"

    - see also: "it's better to be lucky than good"

    - how do we know someone or something is "lucky" or "good"?

    - how do we know a restaurant we visited for the first time is actually good?

    - how do we know we didn't just happen to get their best dish on a good night?

---
# Tractable imperfect information

- How do we estimate models where a person has imperfect information?

- We've done some of this already with dynamic discrete choice models: 

    - people can't see the future

    - instead, have expectations about their future states & preference shocks 
    
    - we compute individuals' expectations according to the $\mathbb{E}\max$ formula

    - we impose a strong assumption on the distribution of $\epsilon$
    
    - That way, we can tractably compute the $\mathbb{E}\max$


---
# How learning works

- Consider a setting where an agent is trying to learn about something, call it $a_i$

- For simplicity, assume $a_i$ is continuous and drawn from CDF $F_a$

- The agent doesn't know the exact value of $a_i$, but has beliefs denoted $\mathbb{E}_t [a_i]$

- He gains additional information about $a_i$ from a noisy signal $S_{it}$

- That is, $S_{it} = a_i + \varepsilon_{it}$ where $\varepsilon_{it}$ is pure noise

- The agent updates his beliefs to $\mathbb{E}_{t+1} [a_i]$ by incorporating the new information in $S_{it}$

- This process repeats itself in each period where $S_{it}$ is received


---
# A little bit more math

- If $\varepsilon_{it}$ is pure noise, then it is independent of $a_i$ for all $t$

- Assume WLOG that $\mathbb{E}(a_{i})=0$ and $\mathbb{E}(\varepsilon_{it})=0$ for all $t$

- Then we can decompose the variance of the signal $S_{it}$

\begin{align*}
\mathbb{V}(S_{it}) &= \mathbb{V}(a_{i}) + \mathbb{V}(\varepsilon_{it})\\
&= \sigma^2_a + \sigma^2_{\varepsilon}
\end{align*}

- The .hi[signal-to-noise ratio (SNR)] is defined as
\begin{align*}
\frac{\mathbb{V}(S_{it})}{\mathbb{V}(\varepsilon_{it})} &= \frac{\sigma^2_a + \sigma^2_{\varepsilon}}{\sigma^2_{\varepsilon}}\\
\end{align*}

- This ratio measures the quality of the signal (bigger is better)

---
# Semantic detail

- Some people call the signal the "meaningful input" rather than the input + noise

- In this case, the SNR would be
\begin{align*}
\frac{\mathbb{V}(a_{i})}{\mathbb{V}(\varepsilon_{it})} &= \frac{\sigma^2_a}{\sigma^2_{\varepsilon}}\\
\end{align*}

- I couldn't find a consensus on this, so just make sure you keep track of things

---
# Connection to factor models

- In factor models, we have $J$ correlated noisy measurements $M_j$

- We try to separate the "factor" from $M$ using correlation across the $M_j$'s

- Factor models also involve a composite error term (can label it $a_i + \varepsilon_{ij}$)

- .hi[Difference:] True value of the factor is known to the individual but not the researcher

- In learning models, the factor is unknown to both the researcher and the individual

- Learning models require panel data

---
# Learning models and factor models

- Rather than being substitutes, these two types of models are .hi[complements]

- Factor models recognize that agents might possess some private information

- Learning models underscore the potential importance of unknown information

- If we ignored private information, that might distort what we call "learning"


---
# Bayesian updating of beliefs

- How exactly do agents update their beliefs given new information in $S_{it}$?

- The simplest way to handle this is to assume .hi[Bayesian updating]

- As the name implies, this comes from Bayes' rule

- Given a .hi[prior] mean and variance belief $\mathbb{E}_t[a_i]$ and $\mathbb{V}_t[a_i]$, agents update according to
\begin{align*}
\mathbb{E}_{t+1}[a_i] &= \mathbb{E}_t[a_i]\frac{\sigma^2_\varepsilon}{\sigma^2_\varepsilon + \mathbb{V}_t[a_i]} + S_{it}\frac{\mathbb{V}_t[a_i]}{\sigma^2_\varepsilon + \mathbb{V}_t[a_i]} \\
\mathbb{V}_{t+1}[a_i] &= \mathbb{V}_t[a_i] \frac{\sigma^2_\varepsilon}{\sigma^2_\varepsilon + \mathbb{V}_t[a_i]}
\end{align*}

- $\mathbb{E}_{t+1}[a_i]$ and $\mathbb{V}_{t+1}[a_i]$ are referred to as the .hi[posterior] beliefs

---
# Properties of Bayesian Learning

1. $\mathbb{V}_{t+1}[a_i]>0$ for all $t$
 
    - One is never completely certain of what he has learned

2. If $\sigma^2_a>0$ then $\frac{\partial\mathbb{V}_{t+1}[a_i]}{\partial t}<0$

    - As additional signals are received, uncertainty of beliefs goes down

3. If $\sigma^2_a>0$ then $\lim_{t\rightarrow\infty}\mathbb{V}_{t+1}[a_i] = 0$

    - In the limit, uncertainty of beliefs vanishes

4. The .hi[speed of learning] is dictated by the signal-to-noise ratio

- These properties may not be desirable, but they are intrinsic to Bayesianism

---
# Non-Bayesian updating

- Bayesian updating is so popular because the math works out nicely

- This is because of what is known as a .hi[conjugate prior] (read Wikipedia)

- If we assume a different kind of updating, the math could get ugly real quick

- But a lot of properties of Bayesian updating make sense (e.g. $\lim_{t\rightarrow\infty}\mathbb{V}_{t+1}[a_i] = 0$)

- Moreover, we often don't know people's beliefs

- If we had detailed data on people's beliefs, that would allow us to be more flexible

    - Sort of like how getting stated preference data aids estimation of choice models

---
# Other considerations

- Things get more complicated if the signal is not continuous

    - Naturally, a discrete signal will provide less information
    
    - e.g. Pass/Fail on an exam, versus a 0-100 score

- Another complication is if the signal is selected

    - For example, I only see a wage signal if I have a job
    
    - In this case, we need a choice model to resolve the sample selection problem
    
    - We'll talk about this towards the end of today's class

---
# The Kalman filter

- The .hi[Kalman filter] is a generalization of Bayesian updating of a learning model

- Most common [application](http://greg.czerniak.info/guides/kalman1/): remote sensing of aircraft/spacecraft

    - Any given sensor sends back a "noisy" signal about exact location
    
    - Multiple sensors acting in sequence can provide more reliable location info
    
- Another cool example: estimating the $R_0$ of SARS-Cov-2

    -  `r Citet(biblio,"kalmanCovidR0")`
    
    - They created a neat continuously updated [dashboard](http://trackingr-env.eba-9muars8y.us-east-2.elasticbeanstalk.com/)
    
    - $R_0$ tends to 1 in equilibrium, as predicted by [Joshua Gans](https://joshuagans.substack.com/p/why-r-tends-towards-1)

---
# Multidimensional learning

- What would Bayesian updating look like if $a_i$ were a vector rather than a scalar?

- Let $A_i$ denote the vector, and suppose its population covariance is $\Delta$

- $\mathbf{S}_{it} = A_i + \boldsymbol\varepsilon_{it}$ is a vector-valued signal

\begin{align*}
\mathbb{E}_{t+1}[A_i]&=(\mathbb{V}^{-1}_{t}[A_i] + \Omega_{it})^{-1}(\mathbb{V}^{-1}_{t}[A_i]\mathbb{E}_{t}[A_i]+\Omega_{it}\mathbf{S}_{it}) \\
\mathbb{V}_{t+1}[A_i]&=(\mathbb{V}^{-1}_{t}[A_i] + \Omega_{it})^{-1}
\end{align*}

- $\Omega_{it}$ is a diagonal matrix with $\frac{1}{\sigma^2_{\varepsilon_j}}$ in the $(j,j)$ element

- Elements of $\Omega$ and $\mathbf{S}$ are set to 0 for signals that aren't received

    - (in other words, not all signals need to be received in every period)

---
# Updating, step by step

- This example will hopefully clarify how updating works

- In period 1, the individual begins with prior beliefs $(\mathbb{E}_1[a_i],\mathbb{V}_1[a_i])$

- Usually, set these to the population values $(0,\sigma^2_a)$ for all individuals

- Then, a signal $S_{i1}$ is received and beliefs are updated according to the formulas:
\begin{align*}
\mathbb{E}_{2}[a_i] &= \underbrace{\mathbb{E}_1[a_i]}_{0}\frac{\sigma^2_\varepsilon}{\sigma^2_\varepsilon + \underbrace{\mathbb{V}_1[a_i]}_{\sigma^2_a}} + S_{i1}\frac{\overbrace{\mathbb{V}_1[a_i]}^{\sigma^2_a}}{\sigma^2_\varepsilon + \underbrace{\mathbb{V}_1[a_i]}_{\sigma^2_a}}\\
& = \frac{S_{i1}\sigma^2_a}{\sigma^2_\varepsilon + \sigma^2_a}
\end{align*}

---
# Updating the variance

- When $S_{i1}$ is received, $i$ updates the variance as follows:

\begin{align*}
\mathbb{V}_{2}[a_i] &= \underbrace{\mathbb{V}_1[a_i]}_{\sigma^2_a} \frac{\sigma^2_\varepsilon}{\sigma^2_\varepsilon + \underbrace{\mathbb{V}_1[a_i]}_{\sigma^2_a}}\\
& = \frac{\sigma^2_a\sigma^2_\varepsilon}{\sigma^2_\varepsilon + \sigma^2_a}
\end{align*}

- It is straightforward to show that $\mathbb{V}_{2}[a_i]<\mathbb{V}_{1}[a_i]$ when $\sigma^2_a>0$

---
# Estimating a simple learning model

- Let's estimate a simple learning model

- Suppose the signals are log wage residuals

- Individuals are trying to ascertain their ability from these residuals

\begin{align*}
\log w_{it} &= X\beta + a_i + \varepsilon_{it}
\end{align*}

- We want to estimate $(\beta,\sigma^2_a,\sigma^2_\varepsilon)$ by maximum likelihood

- We also want to recover each person's beliefs at each point in time

- Let's also compare the results with those from other panel data estimators (FE, RE)

- .hi[Note:] our learning model assumes $a_i \perp X$ so is identical to RE

---
# Estimation code

- In the single linear equation case, estimation is identical to RE (so just use RE)

- For a multidimensional learning case, see [this Github repository](https://github.com/tyleransom/LearningModels)

- Let's estimate the simple learning model from the previous slide

.scroll-box-12[
```{julia,eval=F}
using Random, Statistics, LinearAlgebra, DataFrames, DataFramesMeta, CSV, FixedEffectModels, MixedModels
df = CSV.read("nlswlearn.csv")
dfuse = df[df.ln_wage.!=999,:]
# FE
@show reg(dfuse, @formula(ln_wage ~ 1 + exper*exper + collgrad + race1 + fe(idcode)), Vcov.cluster(:idcode))

# RE
categorical!(dfuse, :idcode)
@show fm1 = fit(MixedModel, @formula(ln_wage ~ 1 + exper*exper + collgrad + race1 + (1|idcode)), dfuse)
# gives σ²_ε = .092187 and σ²_a = 0.106297

# Add columns to data indicating the signal (S_{it}) and prior/posterior mean/variances
sig_eps = .092187
df = @transform(df, signal = :ln_wage .- coef(fm1)[1] .- coef(fm1)[2].*:exper .- coef(fm1)[3]*:collgrad .- coef(fm1)[4].*:race1 .- coef(fm1)[5].*:exper.^2,
                    priorEbelief = zeros(length(:ln_wage)),
                    postrEbelief = zeros(length(:ln_wage)),
                    priorVbelief = 0.106297*ones(length(:ln_wage)),
                    postrVbelief = 0.106297*ones(length(:ln_wage)))
# loop through to apply belief formulas
for i = 1:N
    for t=1:T
        rowt = (i-1)*T+t
        row1 = (i-1)*T+t+1
        if df.ln_wage[rowt]==999 # didn't get a signal this period
            df.signal[rowt] = 0
            df.postrEbelief[rowt] = df.priorEbelief[rowt]
            df.postrVbelief[rowt] = df.priorVbelief[rowt]
        else # these are the formulas from the slides
            df.postrEbelief[rowt] = df.priorEbelief[rowt]*(sig_eps./(sig_eps + df.priorVbelief[rowt])) + df.signal[rowt]*(df.priorVbelief[rowt])./(sig_eps + df.priorVbelief[rowt])
            df.postrVbelief[rowt] = df.priorVbelief[rowt]*(sig_eps./(sig_eps + df.priorVbelief[rowt]))
        end
        if t<T  # set prior in t+1 to be posterior from t except in very last period
            df.priorEbelief[row1] = df.postrEbelief[rowt]
            df.priorVbelief[row1] = df.postrVbelief[rowt]
        end
    end
end
```
]

---
# Looking at the belief updating
.scroll-box-20[
```{julia,eval=F}
│ Row   │ ln_wage │ idcode │ t     │ signal     │ priorEbelief │ postrEbelief │ priorVbelief │ postrVbelief │
├───────┼─────────┼────────┼───────┼────────────┼──────────────┼──────────────┼──────────────┼──────────────┤
│ 1     │ 999.0   │ 1      │ 1     │ 0.0        │ 0.0          │ 0.0          │ 0.106297     │ 0.106297     │
│ 2     │ 999.0   │ 1      │ 2     │ 0.0        │ 0.0          │ 0.0          │ 0.106297     │ 0.106297     │
│ 3     │ 1.45121 │ 1      │ 3     │ 0.071744   │ 0.0          │ 0.0384221    │ 0.106297     │ 0.0493702    │
│ 4     │ 1.02862 │ 1      │ 4     │ -0.436915  │ 0.0384221    │ -0.127359    │ 0.0493702    │ 0.0321516    │
│ 5     │ 1.58998 │ 1      │ 5     │ 0.0476647  │ -0.127359    │ -0.082101    │ 0.0321516    │ 0.0238378    │
│ 6     │ 1.78027 │ 1      │ 6     │ 0.17047    │ -0.082101    │ -0.0302093   │ 0.0238378    │ 0.0189402    │
│ 7     │ 1.77701 │ 1      │ 7     │ 0.167209   │ -0.0302093   │ 0.00343808   │ 0.0189402    │ 0.0157121    │
│ 8     │ 1.77868 │ 1      │ 8     │ 0.168878   │ 0.00343808   │ 0.0275291    │ 0.0157121    │ 0.0134241    │
│ 9     │ 2.49398 │ 1      │ 9     │ 0.825968   │ 0.0275291    │ 0.129018     │ 0.0134241    │ 0.0117178    │
│ 10    │ 2.55172 │ 1      │ 10    │ 0.883707   │ 0.129018     │ 0.214128     │ 0.0117178    │ 0.0103963    │
│ 11    │ 999.0   │ 1      │ 11    │ 0.0        │ 0.214128     │ 0.214128     │ 0.0103963    │ 0.0103963    │
│ 12    │ 2.42026 │ 1      │ 12    │ 0.752253   │ 0.214128     │ 0.268664     │ 0.0103963    │ 0.00934272   │
│ 13    │ 2.61417 │ 1      │ 13    │ 0.946164   │ 0.268664     │ 0.331007     │ 0.00934272   │ 0.008483     │
│ 14    │ 2.53637 │ 1      │ 14    │ 0.868366   │ 0.331007     │ 0.376288     │ 0.008483     │ 0.00776818   │
│ 15    │ 2.46293 │ 1      │ 15    │ 0.746001   │ 0.376288     │ 0.405021     │ 0.00776818   │ 0.00716446   │
├───────┼─────────┼────────┼───────┼────────────┼──────────────┼──────────────┼──────────────┼──────────────┤
│ 16    │ 999.0   │ 2      │ 1     │ 0.0        │ 0.0          │ 0.0          │ 0.106297     │ 0.106297     │
│ 17    │ 999.0   │ 2      │ 2     │ 0.0        │ 0.0          │ 0.0          │ 0.106297     │ 0.106297     │
│ 18    │ 999.0   │ 2      │ 3     │ 0.0        │ 0.0          │ 0.0          │ 0.106297     │ 0.106297     │
│ 19    │ 1.36035 │ 2      │ 4     │ -0.019122  │ 0.0          │ -0.0102407   │ 0.106297     │ 0.0493702    │
│ 20    │ 1.2062  │ 2      │ 5     │ -0.259337  │ -0.0102407   │ -0.0971167   │ 0.0493702    │ 0.0321516    │
```
]


---
# Why learning is important

- Uncertainty and learning can explain some empirical puzzles

    - Why engage in something costly and not finish it? (e.g. college)

- Agents might act differently if they have greater amounts of information

- This means we need to model how beliefs map into actions

- Thus, .hi[learning should be part of a dynamic choice model]

- A persistent question is how can we help agents become more informed?

- Information is valuable, but usually costly to obtain. How can we lower that cost?

---
# Papers that use learning models

- Education:

    - High school dropout `r Citep(biblio,"fu_al2020")`
    
    - College dropout `r Citep(biblio,c("stinebricknerStinebrickner2014jole", "arcidiacono_al2016"))`
    
    - College major choice `r Citep(biblio,c("arcidiacono2004", "stinebricknerStinebrickner2014restud"))`


---
# Papers that use learning models

- Labor:
    
    - Occupational choice `r Citep(biblio,c("miller1984","james2011"))`
    
    - Employee quality `r Citep(biblio,c("farberGibbons1996","altonjiPierret2001"))`

- Family:

    - Marriage match quality `r Citep(biblio,"brien_al2006")`

- IO:

    - Learning about experience goods `r Citep(biblio,c("erdemKeane1996","ackerberg2003"))`

---
# References
.tinier[
```{r refs, echo=FALSE, results="asis"}
PrintBibliography(biblio)
```
]
