---
title: "Lecture 2"
subtitle: "Introduction to Structural Models"
author: Tyler Ransom
date: ECON 6343, University of Oklahoma
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'ou-colors.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

---

```{r, load_refs, include=FALSE, cache=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           hyperlink = FALSE,
           dashed = TRUE)
biblio <- ReadBib("../../References/References.bib", check = FALSE)
#biblio <- ReadBib(system.file("Bib", "biblatexExamples.bib", package = "RefManageR"))
```

# Today and next class

- What is a structural model?

    - Examples

- What steps are required to write a paper that uses a structural model?

    - Computational steps

---

# What is a structural model?

- A .hi[structural model] is an econometric model that relies heavily on economic theory  `r Citep(biblio,"reissWolak2007")`

- The researcher places "structure" (i.e. assumptions) on the model to make estimating the model feasible

- Structure can come either from economic theory or statistics

    - e.g. agents possess private information (economics)
    
    - e.g. wages have a log-normally distributed idiosyncratic component (statistics)

- Structure is needed because data is noisy (either mismeasured or not perfectly in line with the economic model)

---

# Structural parameters

- Structural models assume that the parameters of the model are policy-invariant

- Thus the key parameters tend to be so-called .hi[economic fundamentals]:
    
    - preferences
    
    - production technology
    
    - strategic behavior
    
    - information and expectations
    
    - social interaction (similar to strategic behavior above)

---

# Causality: the goal of econometrics

- As with any econometric endeavor, the goal is to uncover causal relationships

- Causality is crucial because it's the only way to know the effects of a policy

    - e.g. what is the effect of reading to my child on her cognitive development?

    - we can't answer this by simply looking at $Corr(read\,to\,child, cog.\,test\,score)$

    - this correlation is contaminated by omitted variable bias:

    - parents who regularly read to their children likely also regularly do other things favorable to child development

---

# Causality: the goal of econometrics

- Causality requires knowing a .hi[counterfactual]
    
    - what would've been the outcome if everything were the same except the policy?
    
    - this is the notion of _ceteris paribus_ in principles of economics
    
    - theory says quantity demanded will decrease if price increases, _ceteris paribus_
    
    - what is child's test score if everything were the same _except_ regular reading time?

---

# Ways of estimating causal relationships

- .hi[Average Treatment Effect (ATE)] is how we typically quantify a causal effect

- .hi[Non-structural ways:] randomized and quasi-random experiments (RCT, IV, DID, RD)

    * the treatment group and control group are identical except for the policy variable
    
    * thus, $ATE = \overline{y}_{treatment} - \overline{y}_{control}$ where $y$ is the outcome of interest 
    
- .hi[Structural way:] estimate the Data Generating Process (DGP) 

    * once the DGP is known, use it to simulate data in a counterfactual scenario
    
    * $ATE = \overline{y}_{counterfactual} - \overline{y}_{status\,quo}$ 

- These two ways are not mutually exclusive! In fact, they are mutually improving

---

# Reading-to-Children Example: Reduced-form example

- A non-structural (i.e. .hi[reduced-form]) approach would look like the following:

    1. recruit a group of parents to participate in a reading study
    
    2. randomize into "no-read" and "read" groups
    
    3. after some period of time, give the children a cognitive test
    
    4. compare the average scores of children across each of the groups

---

# Reading-to-Children Example: Structural approach
    
- A structural approach would look like the following:

    1. write a model of child skill formation `r Citep(biblio,"cunha_al2010")`
    
    2. gather data on parental and child time use and child test scores
    
    3. estimate the parameters of the child skill formation model
    
    4. use model to simulate counterfactual data (e.g. where reading time is set to 0)
    
    5. compare average scores in counterfactual and status quo

---

# Reading-to-Children Example: Hybrid approach

- A hybrid approach would do the following:

    1. estimate the skill formation parameters
    
    2. leverage randomization to better estimate/validate the model
    
        - e.g. by allowing for identification of a parameter previously not identifiable
    
        - e.g. recover randomization-implied ATE using structural parameter estimates
    
    4. use the validated structural model to explore other counterfactuals

- A great example of this approach is `r Citet(biblio,"delavandeZafar2019")`

---

# Structural vs. Reduced Form Methods

- There is a lot of animosity between structural and reduced-form practitioners

- `r Citet(biblio,"keane2010")` calls reduced-form methods as "atheoretic"

    - others refer to reduced-form or behavioral approaches as "cuteonomics"

- `r Citet(biblio,"harmless2009")` titled their book _Mostly Harmless Econometrics_

    - implying that structural methods include "harmful" econometrics `r Citet(biblio,"lewbel2019zoo")` 

- In fact, .hi[we need both] methodological approaches to answer policy questions

---

# Structural vs. Reduced Form Methods

- `r Citet(biblio,"blundell2010")` outlines three broad concerns determining what method to use: 

1. The nature of the question to be answered

2. The type and quality of data available 

3. The mechanism by which individuals are allocated or receive the policy

    - i.e. random assignment or obeying rules of economic theory (like utility max.)

"Just as an experiment needs to be carefully designed, the identification of a structural economic model needs to be carefully argued."

"Poorly designed quasi-experiments have little to offer, but so to do poorly focused structural estimations."


---

# What is identification? `r Citep(biblio,"lewbel2019zoo")`

- .hi[Identification:] model parameters being _uniquely determined_ from the _observable population_ that generates the data

- identification is never a question about a sample of data

- it is a question about the population from which the sample is drawn

- there are many different terms for identification in econometrics

- but the unifying definition is the one given above

- `r Citet(biblio,"lewbel2019zoo")` lists 33 different terms from the econometrics literature

   (I include all of the terms on the penultimate slide of this deck)

---

# More formal definition `r Citep(biblio,"lewbel2019zoo")`

- Let $\theta$ denote a set of unknown parameters that we would like to learn about, and ideally, estimate 

- e.g. regressor coefficients, average treatment effects, or error distributions 

- identification asks what could be learned about parameters $\theta$ from observable data

- if we knew the population that data are drawn from, would $\theta$ be known? 

- if not, what could be learned about $\theta$?

---

# Why is identification important? `r Citep(biblio,"lewbel2019zoo")`

- The study of identification logically precedes estimation, inference, and testing 

- For $\theta$ to be identified, alternative values of $\theta$ must imply different distributions of the observable data

- If $\theta$ is not identified, then we cannot hope to find a consistent estimator for $\theta$

- More generally, identification failures complicate statistical analyses of models, so recognizing lack of identification, and searching for restrictions that suffice to attain identification, are fundamentally important problems in econometric modeling

---

#  Reduced-form vs. Structural Identification

- How is "identification" used differently in reduced-form vs. structural econometrics?

- In reduced-form econometrics (a.k.a. causal modeling): 

    - Typically talk of an "identification strategy" (i.e. randomization setup)
    
    - Focus is on estimation of treatment effects, not deep parameters
    
    - Relies on randomization from some kind of randomized or natural experiment
    
- In structural econometrics:

    - Typically talk of "establishing identification" (i.e. sufficient variation in data)
    
    - In complex models, can be difficult to do without imposing more assumptions

---

# The Credibility Revolution `r Citep(biblio,"harmless2009")`

- What makes an identification strategy credible?

- Identification means separating selection from treatment

- This is best done when treatment is randomized

- Randomization is also how the natural sciences make discoveries

    - Typically done via controlled experiments

- The closer a reduced-form model is to an RCT, the better

  (Note that controlled experiments are impossible to do with humans)


---

# Examples of Identification Strategies

- Randomized experiments, field experiments, lab experiments

- Instrumental variables, regression discontinuity

- Difference in differences, synthetic control methods

- Matching methods (nearest neighbor, propensity score, ...)

- OLS that does not suffer from omitted variable bias

- These are almost exclusively estimated using linear econometric models

---

# Credible Structural Models

- What makes a structural model credible?

- At the very least, the model should .hi["fit the data"] (i.e. reproduce key patterns)

- But that is usually a low bar to clear, so additional criteria are required

- Results should also "make sense" (i.e. conform to economic theory)

    - e.g. An upward-sloping demand curve would violate this criterion
    
    - or a result that says agents prefer lower income or fewer profits

- Typically requires modeling heterogeneity in preferences or productivity

- Another difficulty: separating preferences from constraints

---

# Structural Methods

- Unlike reduced-form methods, there is not a set "toolkit" of techniques

- Rather, structural modeling is a bit _ad hoc_ or a bit "Wild West"

- Whereas RF methods almost exclusively focus on linear econometric models,

- Structural methods overwhelmingly require use of non-linear econometric models

    - e.g. discrete choice models, fixed point mappings for equilibrium models, etc.
    
- Structural models are typically estimated by GMM or Maximum Likelihood

- Computational know-how helps speed up the process of estimating these models

- These questions will be the focus of this class


---

# Structural papers in various fields of econ

- .hi[Labor:] `r Citet(biblio,"keaneWolpin1997")`, education investment decisions

- .hi[IO:] `r Citet(biblio,"blp1995")`, demand estimation using market-level data

- .hi[Urban:] `r Citet(biblio,"ahlfeldt_al2015")`, estimation of spatial agglomeration

- .hi[Environmental:] `r Citet(biblio,"rudik2020")`, quantify uncertainty in environmental IAMs

- .hi[Public:] `r Citet(biblio,"bayer_al2016")`, dynamic Tiebout sorting model

- .hi[Macro:] all DSGE models

- .hi[International:] `r Citet(biblio,"jin_shen2020")`, coordination of sovereign debt

---

# Internal and External Validity

- .hi[Internal validty] refers to how causal an estimated parameter is

    - "This approach is internally valid" $\Rightarrow$ no selection bias

- .hi[External validity] refers to generalizability of estimates to new contexts

- Typically, RF approaches are very good at internal validity but not at external validity

- On the other hand, if economic agents behave similarly across contexts, structural models can be externally valid

- RF and structural methods used together can improve both internal and external validity

---

# 33 terms for identification

.pull-left[
.smallest[
- Bayesian identification
- causal identification
- essential identification
- eventual identification
- exact identification
- first order identification
- frequentist identification
- generic identification
- global identification
- identification arrangement
- identification at infinity
- identification by construction
- identification of bounds
- ill-posed identification
- irregular identification
- local identification
]
]

.pull-right[
.smallest[
- nearly weak identification
- nonparametric identification
- non-robust identification
- nonstandard weak identification
- overidentification
- parametric identification
- partial identification
- point identification
- sampling identification
- semiparametric identification
- semi-strong identification
- set identification
- strong identification
- structural identification
- thin set identification
- underidentification
- weak identification
]
]

---

# References
.tiny[
```{r refs, echo=FALSE, results="asis"}
#PrintBibliography(biblio, start = 1, end = 2)
PrintBibliography(biblio)
```
]