\documentclass[12pt,english]{article}
\usepackage{mathptmx}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{geometry}
\usepackage{xcolor}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage{amsmath}
\usepackage[authoryear]{natbib}
\usepackage{minted}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=false,bookmarksopen=false,
 breaklinks=false,pdfborder={0 0 0},pdfborderstyle={},backref=false,colorlinks=false]
 {hyperref}
\usepackage{breakurl}

\begin{document}

\title{Problem Set 3}
\author{ECON 6343: Econometrics III\\
Prof. Tyler Ransom\\
University of Oklahoma}
\date{Due: September 15, 9:00 AM}

\maketitle
Directions: Answer all questions. Each student must turn in their own copy, but you may work in groups. Clearly label all answers. Show all of your code. Turn in jl-file(s), output files and writeup via GitHub. Your writeup may simply consist of comments in jl-file(s). If applicable, put the names of all group members at the top of your writeup or jl-file.

You will need to load the following packages:
\begin{itemize}
    \item[~] \texttt{Optim} 
    \item[~] \texttt{HTTP} 
    \item[~] \texttt{GLM} 
    \item[~] \texttt{LinearAlgebra} 
    \item[~] \texttt{Random} 
    \item[~] \texttt{Statistics} 
    \item[~] \texttt{DataFrames} 
    \item[~] \texttt{CSV} 
    \item[~] \texttt{FreqTables}
\end{itemize}

On Github there is a file called \texttt{PS2\_starter.jl} that has the code blocks below already created.

\begin{enumerate}
\item Estimate a multinomial logit (with alternative-specific covariates $Z$) on the following data set:

\begin{minted}[bgcolor=bg]{julia}
using DataFrames
using CSV
using HTTP
url = "https://raw.githubusercontent.com/OU-PhD-Econometrics/fall-2020/
master/ProblemSets/PS3-gev/nlsw88w.csv"
df = CSV.read(HTTP.get(url).body)
X = [df.age df.white df.collgrad]
Z = hcat(df.elnwage1, df.elnwage2, df.elnwage3, df.elnwage4, 
         df.elnwage5, df.elnwage6, df.elnwage7, df.elnwage8)
y = df.occupation
\end{minted}

The choice set represents possible occupations and is structured  as follows.

\begin{enumerate}
    \item[1] Professional/Technical 
    \item[2] Managers/Administrators
    \item[3] Sales                  
    \item[4] Clerical/Unskilled     
    \item[5] Craftsmen              
    \item[6] Operatives             
    \item[7] Transport              
    \item[8] Other                  
\end{enumerate}

\textbf{Hints:} 
\begin{itemize}
    \item Index the parameter vector so that the coefficient on $Z$ is the last element and the coefficients on $X$ are the first set of elements.
    \item You will need to difference the $Z$'s in your likelihood function. 
    \item Normalize $\beta_J = 0$
    \item The formula for the choice probabilities will thus be
    \begin{align*}
        P_{ij} &= \begin{cases} \frac{\exp\left(X_{i}\beta_j + \gamma(Z_{ij}-Z_{iJ})\right)}{1+\sum_{k=1}^{J-1}\exp\left(X_{i}\beta_k + \gamma(Z_{ik}-Z_{iJ})\right)} ,& j=1,\ldots,J-1\\
         \frac{1}{1+\sum_{k=1}^{J-1}\exp\left(X_{i}\beta_k + \gamma(Z_{ik}-Z_{iJ})\right)} ,& j=J
         \end{cases}
    \end{align*}
\end{itemize}

\item Interpret the estimated coefficient $\hat{\gamma}$.

\item Estimate a nested logit with the following nesting structure:
\begin{itemize}
    \item White collar occupations (indexed by $WC$)
    \begin{itemize}
    \item[1] Professional/Technical 
    \item[2] Managers/Administrators
    \item[3] Sales 
    \end{itemize}
    \item Blue collar occupations (indexed by $BC$)
    \begin{itemize}
    \item[4] Clerical/Unskilled     
    \item[5] Craftsmen              
    \item[6] Operatives             
    \item[7] Transport              
    \end{itemize}
    \item Other occupations (indexed by $\text{Other}$)
    \begin{itemize}
    \item[8] Other 
    \end{itemize}
\end{itemize}

Specify the parameters such that there are only nest-level (rather than alternative-level) coefficients. That is, estimate a model with the following parameters:
\begin{itemize}
    \item $\beta_{WC}$
    \item $\beta_{BC}$
    \item $\lambda_{WC}$
    \item $\lambda_{BC}$
    \item $\gamma$
    \item $\beta_{\text{Other}}$ is normalized to 0
    \item The formula for the choice probabilities will thus be
    \begin{align*}
        P_{ij} &= \begin{cases} \frac{\exp\left(\frac{X_{i}\beta_{WC}+\gamma(Z_{ij}-Z_{iJ})}{\lambda_{WC}}\right)\left[\sum_{\ell\in WC}\exp\left(\frac{X_{i}\beta_{WC}+\gamma(Z_{i\ell}-Z_{iJ})}{\lambda_{WC}}\right)\right]^{\lambda_{WC}-1}}{1+\left[\sum_{k\in WC}\exp\left(\frac{X_{i}\beta_{WC}+\gamma(Z_{ik}-Z_{iJ})}{\lambda_{WC}}\right)\right]^{\lambda_{WC}}+\left[\sum_{m\in BC}\exp\left(\frac{X_{i}\beta_{BC}+\gamma(Z_{im}-Z_{iJ})}{\lambda_{BC}}\right)\right]^{\lambda_{BC}}} ,& j\in WC\\
        \frac{\exp\left(\frac{X_{i}\beta_{BC}+\gamma(Z_{ij}-Z_{iJ})}{\lambda_{BC}}\right)\left[\sum_{\ell\in BC}\exp\left(\frac{X_{i}\beta_{BC}+\gamma(Z_{i\ell}-Z_{iJ})}{\lambda_{BC}}\right)\right]^{\lambda_{BC}-1}}{1+\left[\sum_{k\in WC}\exp\left(\frac{X_{i}\beta_{WC}+\gamma(Z_{ik}-Z_{iJ})}{\lambda_{WC}}\right)\right]^{\lambda_{WC}}+\left[\sum_{m\in BC}\exp\left(\frac{X_{i}\beta_{BC}+\gamma(Z_{im}-Z_{iJ})}{\lambda_{BC}}\right)\right]^{\lambda_{BC}}} ,& j\in BC\\
         \frac{1}{1+\left[\sum_{k\in WC}\exp\left(\frac{X_{i}\beta_{WC}+\gamma(Z_{ik}-Z_{iJ})}{\lambda_{WC}}\right)\right]^{\lambda_{WC}}+\left[\sum_{m\in BC}\exp\left(\frac{X_{i}\beta_{BC}+\gamma(Z_{im}-Z_{iJ})}{\lambda_{BC}}\right)\right]^{\lambda_{BC}}} ,& j=J
         \end{cases}
    \end{align*}

\end{itemize}


\item Wrap all of your code above into a function and then call that function at the very bottom of your script. Make sure you add \texttt{println()} statements after obtaining each set of estimates so that you can read them.
\end{enumerate}
\end{document}