\documentclass[12pt,english]{article}
\usepackage{mathptmx}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{geometry}
\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\definecolor{byublue}     {RGB}{0.  ,30. ,76. }
\definecolor{darkblue}    {RGB}{0.  ,0.  ,139.}
\definecolor{dukeblue}    {RGB}{0.  ,0.  ,156.}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage{amsmath}
\usepackage[authoryear]{natbib}
\usepackage{minted}
\usepackage{mathtools}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\usepackage[backref=page]{hyperref}                                              % Always add hyperref (almost) last
\hypersetup{unicode=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=3,
 breaklinks=true,pdfborder={0 0 0},colorlinks,citecolor=darkblue,filecolor=darkblue,linkcolor=darkblue,urlcolor=darkblue,pagebackref=true}
\usepackage[all]{hypcap}                                            % Links point to top of image, builds on hyperref
\usepackage{breakurl}

\begin{document}

\title{Referee Report Assignment}
\author{ECON 6343: Econometrics III\\
Prof. Tyler Ransom\\
University of Oklahoma}
\date{Due: December 10, 9:00 AM}

\maketitle
Directions: 

\begin{enumerate}
    \item Read \href{https://www.aeaweb.org/content/file?id=222}{this article} on the American Economic Association's website. 
    \item Select a paper to review that is related in some way to the topics we've covered this semester. The paper should not be published. If it has been published, then you should find the earliest pre-print version you can and use that to base your review on. Note that Google Scholar does an excellent job of keeping track of multiple versions of papers. Many well-published papers were often released in a working paper or discussion paper series. These working paper or pre-print series can also be a source for you.
    \item Following the guidelines of the article referenced above, prepare a cover letter addressed to me, the (fake) editor. Also prepare a referee report that could be shared with the author(s).
    \item Turn in your documents as PDFs to the appropriate area on Canvas.
\end{enumerate}

\end{document}