---
title: "Lecture 9"
subtitle: "Simulated Method of Moments: Another Method of Structural Estimation"
author: Tyler Ransom
date: ECON 6343, University of Oklahoma
output:
  xaringan::moon_reader:
    includes:
        in_header: "09slides_files/mathjax-equation-numbers.html"
    css: ['default', 'metropolis', 'metropolis-fonts', 'ou-colors.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

---

```{r, load_refs, include=FALSE, cache=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           style = "markdown",
           hyperlink = FALSE,
           dashed = TRUE)
biblio <- ReadBib("../../References/References.bib", check = FALSE)
#biblio <- ReadBib(system.file("Bib", "biblatexExamples.bib", package = "RefManageR"))
```

# Plan for the Day

1. Review Method of Moments and GMM

2. Introduce simulated method of moments (SMM)

3. Walk through how to do SMM in Julia

4. Discuss indirect inference


---
# Generalized Method of Moments (GMM)

- GMM is a fundamental concept taught in graduate-level econometrics

- It is very popular because it nests many common econometric estimators:
    - OLS
    - IV and 2SLS
    - Nonlinear least squares (NLLS)
    - MLE (e.g. probit, logit)

- There's a great overview video [here](https://www.youtube.com/watch?v=U7Ylm187hYA)

---
# Method of Moments

- We can use method of moments to estimate a model's parameters

- Consider a simple regression model

\begin{align*}
y &= \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \varepsilon
\end{align*}

- Assume $\mathbb{E}[\varepsilon \vert \mathbb{x}] = 0$ (conditional independence)

- Then we can form a system of 3 equations and 3 unknowns


---
# OLS Population Moment Conditions

- If we write out the OLS moment conditions, we get

\begin{align*}
\mathbb{E}[\varepsilon]     &= 0\\
\mathbb{E}[\varepsilon' x_1] &= 0\\
\mathbb{E}[\varepsilon' x_2] &= 0\\
\end{align*}

- Rewriting in terms of our parameters of interest $(\beta_0,\beta_1,\beta_2)$:

\begin{align*}
\mathbb{E}[(y - \beta_0 - \beta_1 x_1 - \beta_2 x_2)]     &= 0\\
\mathbb{E}[(y - \beta_0 - \beta_1 x_1 - \beta_2 x_2)' x_1] &= 0\\
\mathbb{E}[(y - \beta_0 - \beta_1 x_1 - \beta_2 x_2)' x_2] &= 0\\
\end{align*}

---
# OLS Sample Moment Conditions

- We then need to adjust the previous formula to work with sample analogs:
\begin{align*}
g\left(\boldsymbol \beta\right) &=\begin{cases}
\frac{1}{N}\sum_{i=1}^N(y_i - \beta_0 - \beta_1 x_{i1} - \beta_2 x_{i2})         &= 0\\
\frac{1}{N}\sum_{i=1}^N(y_i - \beta_0 - \beta_1 x_{i1} - \beta_2 x_{i2})' x_{i1} &= 0\\
\frac{1}{N}\sum_{i=1}^N(y_i - \beta_0 - \beta_1 x_{i1} - \beta_2 x_{i2})' x_{i2} &= 0\end{cases}
\end{align*}

- We can estimate this by exactly-identified GMM using the objective function

\begin{align*}
\hat{\boldsymbol \beta} &= \arg \min_{\boldsymbol \beta} J\left(\boldsymbol \beta\right)
\end{align*}
where
\begin{align*}
J\left(\boldsymbol \beta\right) &= N g\left(\boldsymbol \beta\right)' g\left(\boldsymbol \beta\right)
\end{align*}

---
# GMM with more moment conditions than parameters

- The solution to the obj fn on the pvs slide has a closed form for OLS: $(X'X)^{-1}X'y$

- In cases with more moment conditions than parameters, we need to weight
\begin{align*}
\hat{\boldsymbol \beta} &= \arg \min_{\boldsymbol \beta} J\left(\boldsymbol \beta, \hat{\mathbf{W}}\right)
\end{align*}
where
\begin{align*}
J\left(\boldsymbol \beta\right) &= N g\left(\boldsymbol \beta\right)' \hat{\mathbf{W}}(\boldsymbol \beta) g\left(\boldsymbol \beta\right)
\end{align*}

- There is a ton of econometric theory about the optimal weighting matrix $\hat{\mathbf{W}}$

- As well as the asymptotic properties of the GMM estimator (spoiler: they're good)

---
# GMM as OLS

- Another example is OLS posed a different way

- Previously, we solved $K$ equations of $\mathbb{E}\left[\varepsilon'X_k\right]=0$ and $\mathbb{E}\left[\varepsilon\right]=0$

- We could instead simply try to match $y$ to $X\beta$ for every observation

- In this case, $g = y-X\beta$

- There are $N$ moment conditions and $K+1$ parameters to be estimated

- Use the $N\times N$ Identity matrix for $\mathbf{W}$ and this is precisely OLS

- In my experience, this approach has better computational properties than the "classical" approach

---
# Binary Logit Sample Moment Conditions

- The "classical" approach to the moment conditions for the binary logit model is:
\begin{align*}
g\left(\boldsymbol \beta\right) &=\begin{cases}
\frac{1}{N}\sum_{i=1}^N\left[y_i - \frac{\exp\left(\beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}\right)}{1+\exp\left(\beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}\right)}\right]         &= 0\\
\frac{1}{N}\sum_{i=1}^N\left[y_i - \frac{\exp\left(\beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}\right)}{1+\exp\left(\beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}\right)}\right]' x_{i1} &= 0\\
\frac{1}{N}\sum_{i=1}^N\left[y_i - \frac{\exp\left(\beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}\right)}{1+\exp\left(\beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}\right)}\right]' x_{i2} &= 0\end{cases}
\end{align*}
where $y_i \in \left\{0,1\right\}$ 

- With the same formula for $J$ as in the OLS case (or any other case)

- Under the alternative approach, use $g = y - P$ where $y\in \left\{0,1\right\}$ and $P\in\left[0,1\right]$

- Again, there are $N$ moment conditions and $K+1$ parameters to be estimated

---
# Coding example: Estimating binary logit by GMM

- We can estimate the binary logit model by GMM as follows:

``` {julia, eval=F}
using Optim, LinearAlgebra
function logit_gmm(α, X, y)
    P = exp.(X*α)./(1 .+ exp.(X*α))
    g = y .- P
    J = g'*I*g
    return J
end
α̂_optim = optimize(a -> logit_gmm(a, X, y), rand(size(X,2)), LBFGS(), Optim.Options(g_tol=1e-8, iterations=100_000))
println(α̂_optim.minimizer)
```

- This gives estimates that are quite close to (but not identical to) MLE

- With $\mathbf{W} = \mathbf{I}$, this objective function is identical to nonlinear least squares (NLLS)



---
# Usefulness of simulation

- As we showed in PS4, we can sometimes use simulation to compute integrals

- Another alternative is to use quadrature to compute the integral

- In the simulation case, we took draws from the mixture distribution of a mixed logit

- More generally, we can estimate highly complex models using simulation methods

- In some cases, simulation is the _only_ option; everything else is intractable
    - Quadrature typically only works with very low-dimensional integrals

---
# Simulation methods

`r Citet(biblio,"train2009")` mentions three different types of simulation-based methods:

1. .hi[Simulated Maximum Likelihood] (a.k.a. Maximum Simulated Likelihood)

2. .hi[Simulated Method of Moments] (a.k.a. Method of Simulated Moments)

3. .hi[Method of Simulated Scores]

- What I asked you to code up in PS4 was basically SML

- Today we'll talk mostly about SMM

- We won't cover Method of Simulated Scores


---
# Simulated Method of Moments

- As the name would imply, SMM is a simulated version of GMM

- The difference: SMM uses moments from simulated data

- The objective is then to make simulated and actual data match

- See `r Citet(biblio,"mcfadden1989")` and `r Citet(biblio,"evans2018")` for more details

- `r Citet(biblio,"evans2018")` includes a Python coding example

- Notes by [Jason DeBacker](https://www.jasondebacker.com/classes/Lecture10_Notes_SMM.pdf), [Eric Sims](https://www3.nd.edu/~esims1/advanced_topics.pdf) and [Colin Cameron](http://cameron.econ.ucdavis.edu/mmabook/transparencies/ct06_gmm.pdf) are also helpful


---
# Pros of SMM

- Can estimate models with $P$'s that don't have a closed form, like probit `r Citep(biblio,"chintagunta1992")`

- Can estimate other models that would otherwise be intractable
    - e.g. dynamic models with high-dimensional integrals
    
- Or micro-models based only on aggregated data

- Coding for simulating the model is already done! Can dive right into counterfactuals

- It's straightforward to interpret the moments and know that model is fitting these

- Also easier to compare with reduced-form evidence

---
# Cons of SMM

- Much more computationally intensive than GMM

- Loss of (statistical) efficiency, relative to MLE (i.e. larger SE's)

- For me personally, it's not always clear which moments to select
    - this can feel a bit _ad hoc_

---
# SMM in Julia

- Once we know the objective fn, we can program any estimator we please

- Let's consider how to estimate a simple linear regression model

\begin{align*}
y &= X\beta + \varepsilon\\
\varepsilon&\sim N(0,\sigma^2)
\end{align*}

- $y$ and $X$ are data, and we want to estimate $\beta$ and $\sigma$

- .hi[Note:] here we need to make an assumption about what the DGP looks like

- This means making the strong assumption that $\varepsilon\sim N(0,\sigma^2)$


---
# Estimation steps

- For .hi[each guess] of $\theta = [\beta', \sigma]'$ we do the following:

    - Compute data moments
    
    - Draw $N$ $\varepsilon$'s $D$ times (typically $D>1000$)
    
    - For each draw, compute $y$ from the model equation (call it $\tilde{y}$) given $\theta$
    
    - Compute model moments using $\tilde{y}$ (same as data moments with $y$)
    
    - Model moments are averaged across all $D$ draws

    - Update objective function value given values of data and avg'd model moments

---
# SMM in Julia
.scroll-box-12[
```{julia,eval=F}
function ols_smm(θ, X, y, D)
    K = size(X,2)
    N = size(y,1)
    β = θ[1:end-1]
    σ = θ[end]
    if length(β)==1
        β = β[1]
    end
    # N+1 moments in both model and data
    gmodel = zeros(N+1,D)
    
    # data moments are just the y vector itself
    # and the variance of the y vector
    gdata  = vcat(y,var(y))

    #### !!!!!!!!!!!!!!!!!!!!!!!!!!!!! ####
    # This is critical!                   #
    Random.seed!(1234)                    #
    # You must always use the same ε draw #
    # for every guess of θ!               #
    #### !!!!!!!!!!!!!!!!!!!!!!!!!!!!!! ###
    # simulated model moments
    for d=1:D
        ε = σ*randn(N)
        ỹ = X*β .+ ε
        gmodel[1:end-1,d] = ỹ
        gmodel[  end  ,d] = var(ỹ)
    end

    # criterion function
    err = vec(gdata .- mean(gmodel; dims=2))

    # weighting matrix is the identity matrix

    # minimize weighted difference between data and moments
    J = err'*I*err
    return J
end
```
]

- Data moments to match: $\left\{y_i, i=1,\ldots,N;\widehat{V}(y)\right\}$

- Model moments to match: $\left\{\tilde{y}_i, i=1,\ldots,N;\widehat{V}(\tilde{y})\right\}$

---
# SMM optimization

- We can optimize the objective function with any optimizer we'd like

- In general, the SMM objective function may be poorly behaved (i.e. local optima)

- So you may need to employ tactics to find the global optimum:
    - use LBFGS from many different starting values
    - use Simulated Annealing or Particle Swarm
    - (these are algorithms designed to find global optima)
    
- But SMM should be well behaved for simple problems (like OLS)

- .hi[Always remember:] Must use .hi[same] draw of $\varepsilon$ in every optimizer iteration!

---
# SMM.jl

- SMM is so common, that others have already implemented it
    - And probably in a more computationally efficient manner!

- One such package is `SMM.jl`, written by [Florian Oswald](https://floswald.github.io/) (Sciences Po)
    - This package allows for parallelization, which can speed up estimation time
    
    - It also uses a Bayesian Markov Chain Monte Carlo algorithm known as BGP
    
    - "BGP" comes from `r Citet(biblio,"bgp2013")`
    
    - I am still learning this package but there are some examples

---
# SMM.jl example

- Let's estimate the following model using `SMM.jl`

\begin{align*}
Y_1 &= \beta_{01} + \varepsilon_{1}\\
Y_2 &= \beta_{02} + \varepsilon_{2}
\end{align*}
where $\mathbf{\varepsilon} \sim MVN\left(\mathbf{0},I\right)$. Thus, the $\beta$'s constitute the means of each MVN dimension.

- The code to do this is included in the examples of `SMM.jl` with $(\beta_{01},\beta_{02}) = (-1,1)$

```{julia, eval=FALSE}
using SMM, DataFrames
MA = SMM.parallelNormal() # Note: this line may take up to 5 minutes to execute
dc = SMM.history(MA.chains[1])
dc = dc[dc[:accepted].==true, :]
println(describe(dc))
```

- You can then verify that the `mean` column for `p1` and `p2` is close to -1 and 1.

---
# Indirect inference `r Citep(biblio,"smithPalgrave2008")`

- So far today we've only talked about matching model moments to data

- Logic: if the model matches the data, then it is a reasonable model

- Another alternative is known as .hi[indirect inference]

- In this case, we use an .hi[auxiliary model]

---

# Indirect Inference (Cont'd)

- The auxiliary model doesn't need to accurately describe the DGP

- It simply acts a lens through which to view the world

- .hi[Objective:] minimize the parameters of the economic model such that

- real-world data = simulated data .hi[through the lens of the auxiliary model]


---

# Example: Economic Model

- Consider a simple macro model with two simultaneous equations:

\begin{align*}
C_t &= \beta Y_t + u_t\\
Y_t &= C_t + X_t
\end{align*}

- $C_t$ (consumption) and $Y_t$ (income) are endogenous

- $X_t$ (non-consumption expenditure) is exogenous

- $u_t \overset{iid}{\sim}N(0,\sigma^2)$ 

- Supposing we know the value of $\sigma^2$, then $\beta$ is the lone parameter in the model

---
# Example: Auxiliary model

- We don't need to use indirect inference to estimate $\beta$, but we can

- Suppose our auxiliary model is

\begin{align*}
C_t &= \theta X_t + e_t\\
e_t &\sim N(0,s^2)
\end{align*}
where again the variance $s^2$ is known

- We can estimate $\theta$ by OLS or MLE

- But how does that help us estimate $\beta$?

- We need to find the mapping between $\beta$ and $\theta$

---
# Example: Finding the mapping

- Let's apply some algebra to the first system of equations. Substituting $Y_t$ gives

\begin{align*}
C_t &= \beta(C_t+X_t)+u_t\\
C_t &= \frac{\beta}{1-\beta}X_t + \frac{1}{1-\beta}u_t \\
&\Rightarrow \theta = \frac{\beta}{1-\beta} \\
&\Rightarrow \beta = \frac{\theta}{1+\theta}
\end{align*}

- We know we can easily estimate $\theta$ by OLS

- Then we can recover $\hat{\beta}$ by evaluating $\frac{\hat{\theta}}{1+\hat{\theta}}$

- We worked backwards from the auxiliary model to get estimates of the main model


---

# References
.smallest[
```{r refs, echo=FALSE, results="asis"}
#PrintBibliography(biblio, start = 1, end = 2)
PrintBibliography(biblio)
```
]
